Workday sleep, total cholestrol and the modifying effect of age on systolic blood pressure: An analysis of the 2015-2018 non-obese NHANES population

### 2023-12-11 manuscript/paper.Rmd

---
title: "Weekday sleep, total cholestrol and the modifying effect of age on systolic blood pressure: An analysis of the 2015-2018 non-obese NHANES population"
thanks: "Replication files are available on the author's Github account (http://github.com/okutse/sleepBP). **Current version**: `r format(Sys.time(), '%B %d, %Y')`"
#subtitle: "Short title goes here!"
author:
- Amos Okutse, Brown University
- Monic Colon-Vargas, Brown University
abstract: "**Introduction:** Lorem ipsum dolor sit amet, consectetur adipiscing elit. Donec sit amet libero justo. Pellentesque eget nibh ex. Aliquam tincidunt egestas lectus id ullamcorper.

@@

Sleep duration on workdays was evaluated by the questionnaire with the
following questions: "Number of hours usually sleep on weekdays or
workdays". We then categorized this variable into three groups, that is,
\< 6 h, 6--8 h, ≥ 8 h respectively, and used \< 6 h as the reference
group in our analysis. We also explored the association between total
cholestrol (in mmol/L) level on systolic blood pressure including
potential effect modification by other factors.

@@

```{=html}
<!-- Sleep duration on workdays was evaluated by the questionnaire with the
following questions: "Number of hours usually sleep on weekdays or
workdays". We then categorized this variable into three groups, that is,
\< 6 h, 6--8 h, ≥ 8 h respectively, and used \< 6 h as the reference
group in our analysis. -->
```
Our analyses included the following as covariates: Race divided into
four groups as Mexican American, white, black and other race. Alcohol
consumption was grouped into drinking, no drinking, not recorded.
Smoking status as smoking, not smoking, not recorded. Diabetes was
defined as yes, no, borderline, or not recorded. Hypertension was
defined as yes, no, or not recorded. Snorting was defined as yes, no,
and not recorded. US citizenship status was defined as citizen by birth
or naturalization, don't know, not a citizen, or refused to answer.
Education level was grouped into four categories including graduate
studies, high school, less than grade 12 or some college. Additional
covariates included marital status, gender, age, albumin,
creatinine,hemoglobin, total cholesterol (TC), aspertate
aminotransferase (AST), high-density lipoprotein (HDL), and body mass
index (BMI). Details about these variables can be found at
<https://wwwn.cdc.gov/nchs/nhanes/search/default.aspx>. Age, albumin,
creatinine, hemoglobin, TC, AST, HDL, and BMI were analyzed as
continuous variables whereas gender, alcohol consumption, diabetes,
smoking, race, hypertension, and snorting were analyzed as categorical
variables. Table \@ref{tab:tabone} highlights the variable names and
descriptions as utilized in this study.

@@

\caption{\label{tab:tabone}Variable descriptions}
\centering
\begin{tabular}[t]{>{\raggedright\arraybackslash}p{2cm}>{\raggedright\arraybackslash}p{2cm}>{\raggedright\arraybackslash}p{8cm}}
\toprule
\textbf{Variable} & \textbf{Name} & \textbf{Description}\\
\midrule
\textbf{SEQN} & \textbf{sequence number} & Respondent number\\
\textbf{SDMVPSU} & \textbf{psu} & Masked variance unit pseudo-PSU variable for variance estimation\\
\textbf{WTINT2YR} & \textbf{weights} & Full sample 2-year interview weights\\
\textbf{SDMVSTRA} & \textbf{strata} & Masked variance unit pseudo-stratum variable for variance estimation\\
\textbf{RIAGENDR} & \textbf{gender} & Respondent's number\\
\addlinespace
\textbf{RIDAGEYR} & \textbf{age (yrs)} & Respondent's age in years\\
\textbf{DMDMARTL} & \textbf{marital status} & Marital status\\
\textbf{INDFMIN2} & \textbf{income category} & Total family income (reported as a range value in dollars)\\
\textbf{RIDRETH3} & \textbf{race} & Recode of reported race and Hispanic origin information, with Non-Hispanic Asian Category\\
\textbf{DMDHHSZA} & \textbf{children <5} & Number of children aged 5 years or younger in the household\\
\addlinespace
\textbf{DMDEDUC2} & \textbf{education level} & What is the highest grade or level of school\{you have/SP has\} completed or the highest degree \{you have/s/he has\} received?\\
\textbf{DMDCITZN} & \textbf{citizenship status} & \{Are you/Is SP\} a citizen of the United States? [Information about citizenship is being collected by the U.S. Public Health Serv\\
\textbf{SLD012} & \textbf{sleep} & Number of hours usually sleep on weekdays or workdays\\
\textbf{BMXBMI} & \textbf{bmi} & Body mass index\\
\textbf{ALQ121} & \textbf{alcohol use} & In the past 12 months, how often did you drink any type of alcoholic beverage?\\
\addlinespace
\textbf{LBDHDDSI} & \textbf{hdl} & High density lipoprotein\\
\textbf{LBDSALSI} & \textbf{albumin} & Albumin (g/L)\\
\textbf{DIQ010} & \textbf{diabetes} & Have you ever been told by a doctor or health professional that you have diabetes or sugar diabetes?\\
\textbf{BPQ020} & \textbf{hypertension} & \{Have you/Has SP\} ever been told by a doctor or other health professional that \{you/s/he\} had hypertension, also called high blood pressure?\\
\textbf{URXCRS} & \textbf{creatinine} & Creatinine, urine (umol/L)\\
\addlinespace
\textbf{SLQ040} & \textbf{snort} & In the past 12 months, how often did \{you/SP\} snort, gasp, or stop breathing while \{you were/s/he was\} asleep?\\
\textbf{LBDTCSI} & \textbf{total cholestrol levels} & Total cholesterol (mmol/L)\\
\textbf{LBXHGB} & \textbf{hemoglobin} & Hemoglobin (g/dL)\\
\textbf{LBXSASSI} & \textbf{AST} & Aspertate aminotransferase\\
\textbf{SMQ040} & \textbf{smoke} & Do you now smoke cigarettes?\\
\addlinespace
\textbf{BPXDI} & \textbf{blood pressure} & Systolic and diastolic blood pressure taken as the average of the first three measurements\\
\bottomrule
\end{tabular}
\end{table}
```
```{r tabone, echo=FALSE, eval=FALSE}
tab.one <- vars %>% 
  kable(format = "latex", caption = "Variable descriptions", booktabs = TRUE) %>% 
  kable_styling(latex_options = c("HOLD_position")) %>%
  row_spec(c(0), bold = TRUE) %>%
  column_spec(c(1, 2), bold = TRUE, width = "2cm") %>%
  column_spec(3, width = "8cm")
tab.one
```

@@

Univariate regression
analysis results exploring the association between the exposures,
selected potential confounding variables showed substantial associations between BMI ($\beta = 0.95$ mmHg; 95%CI [0.60 - 1.3]; p<0.05), TC ($\beta = 3.6$ mmHg; 95%CI [2.9 - 4.3]; p<0.05), albumin ($\beta = -0.48$ mmHg; 95% CI [-0.68mmHg - -0.27]; p < 0.05), AST ($\beta = 0.12$ mmHg; 95%CI [0.07 - 0.18]; p <0.05), and age (years) ($\beta = 0.49$ mmHg; 95% CI [0.45 - 0.52]; p <0.05) and systolic blood pressure. Weekday sleep duration between 6 and 8 hours was associated with a 4.0 mmHg (95% CI [-6.8 - -1.2]) decrease in the average systolic blood pressure compared to 6 hours of sleep. On the other hand, more than 8 hours of sleep was associated with a decrease of 3.1 mmHg in systolic blood compared to 6 hours of sleep. Substantial associations were noted between citizenship, education, number of children below five years, gender, marital status, race, smoking status, snorting, alcohol, and income.

@@

Table \@ref(tab:multreg) shows the coefficients of sleep, age, total
cholesterol and the interaction between total cholesterol and age

@@

\begin{table}[H]
\caption{\label{tab:multreg} Sleep, Age and Cholesterol Coefficients for Model 5}
\centering
\begin{tabular}[t]{lrrr}
\toprule
  \textbf{Variable} & \textbf{Estimate} & \textbf{95\% CI}  & \textbf{p-value} \\
\midrule
\textbf{Intercept} & 102.55 &(88.42,116.69) & $<0.001$\\
\textbf{Sleep} & 0.75 & (0.50, 1.00) & 0.002 \\
$<6$hrs &  &  &  \\
$>8$hrs & -1.4 & (-3.7,0.92) & 0.2\\
$6-8$hrs & -3.0 & (-5.3,-0.67) &  0.012\\
\textbf{Age} & -0.04& (-0.22, 0.13)	& 0.6 \\
\textbf{TC} & -3.5	& (-5.2, -1.8) & <0.001 \\
\textbf{TC*Age} & 0.11	& (0.07, 0.14) & <0.001 \\
\bottomrule
\end{tabular}
\end{table}
```

@@

>>>>

---
title: "Workday sleep, total cholestrol and the modifying effect of age on systolic blood pressure: An analysis of the 2015-2018 non-obese NHANES population"
thanks: "Replication files are available on the author's Github account (http://github.com/okutse/sleepBP). **Current version**: `r format(Sys.time(), '%B %d, %Y')`"
#subtitle: "Short title goes here!"
author:
- Amos Okutse, Brown University
- Monic Colon-Vargas, Brown University
abstract: "**Introduction:** Lorem ipsum dolor sit amet, consectetur adipiscing elit. Donec sit amet libero justo. Pellentesque eget nibh ex. Aliquam tincidunt egestas lectus id ullamcorper.

@@

Sleep duration on workdays was evaluated by the questionnaire with the
following questions: "Number of hours usually sleep on Workdays or
workdays". We then categorized this variable into three groups, that is,
\< 6 h, 6--8 h, ≥ 8 h respectively, and used \< 6 h as the reference
group in our analysis. We also explored the association between total
cholestrol (in mmol/L) level on systolic blood pressure including
potential effect modification by other factors.

@@

```{=html}
<!-- Sleep duration on workdays was evaluated by the questionnaire with the
following questions: "Number of hours usually sleep on Workdays or
workdays". We then categorized this variable into three groups, that is,
\< 6 h, 6--8 h, ≥ 8 h respectively, and used \< 6 h as the reference
group in our analysis. -->
```
Our analyses included the following as covariates based on previous literature [@su2022association]: Race divided into
four groups as Mexican American, white, black and other race. Alcohol
consumption was grouped into drinking, no drinking, not recorded.
Smoking status as smoking, not smoking, not recorded. Diabetes was
defined as yes, no, borderline, or not recorded. Hypertension was
defined as yes, no, or not recorded. Snorting was defined as yes, no,
and not recorded. US citizenship status was defined as citizen by birth
or naturalization, don't know, not a citizen, or refused to answer.
Education level was grouped into four categories including graduate
studies, high school, less than grade 12 or some college. Additional
covariates included marital status, gender, age, albumin,
creatinine,hemoglobin, total cholesterol (TC), aspertate
aminotransferase (AST), high-density lipoprotein (HDL), and body mass
index (BMI). Details about these variables can be found at
<https://wwwn.cdc.gov/nchs/nhanes/search/default.aspx>. Age, albumin,
creatinine, hemoglobin, TC, AST, HDL, and BMI were analyzed as
continuous variables whereas gender, alcohol consumption, diabetes,
smoking, race, hypertension, and snorting were analyzed as categorical
variables. Table \@ref{tab:tabone} highlights the variable names and
descriptions as utilized in this study.

@@

\caption{\label{tab:tabone}Variable descriptions}
\centering
\begin{tabular}[t]{>{\raggedright\arraybackslash}p{2cm}>{\raggedright\arraybackslash}p{2cm}>{\raggedright\arraybackslash}p{8cm}}
\toprule
\textbf{Variable} & \textbf{Name} & \textbf{Description}\\
\midrule
\textbf{SEQN} & \textbf{sequence number} & Respondent number\\
\textbf{SDMVPSU} & \textbf{psu} & Masked variance unit pseudo-PSU variable for variance estimation\\
\textbf{WTINT2YR} & \textbf{weights} & Full sample 2-year interview weights\\
\textbf{SDMVSTRA} & \textbf{strata} & Masked variance unit pseudo-stratum variable for variance estimation\\
\textbf{RIAGENDR} & \textbf{gender} & Respondent's number\\
\addlinespace
\textbf{RIDAGEYR} & \textbf{age (yrs)} & Respondent's age in years\\
\textbf{DMDMARTL} & \textbf{marital status} & Marital status\\
\textbf{INDFMIN2} & \textbf{income category} & Total family income (reported as a range value in dollars)\\
\textbf{RIDRETH3} & \textbf{race} & Recode of reported race and Hispanic origin information, with Non-Hispanic Asian Category\\
\textbf{DMDHHSZA} & \textbf{children <5} & Number of children aged 5 years or younger in the household\\
\addlinespace
\textbf{DMDEDUC2} & \textbf{education level} & What is the highest grade or level of school\{you have/SP has\} completed or the highest degree \{you have/s/he has\} received?\\
\textbf{DMDCITZN} & \textbf{citizenship status} & \{Are you/Is SP\} a citizen of the United States? [Information about citizenship is being collected by the U.S. Public Health Serv\\
\textbf{SLD012} & \textbf{sleep} & Number of hours usually sleep on Workdays or workdays\\
\textbf{BMXBMI} & \textbf{bmi} & Body mass index\\
\textbf{ALQ121} & \textbf{alcohol use} & In the past 12 months, how often did you drink any type of alcoholic beverage?\\
\addlinespace
\textbf{LBDHDDSI} & \textbf{hdl} & High density lipoprotein\\
\textbf{LBDSALSI} & \textbf{albumin} & Albumin (g/L)\\
\textbf{DIQ010} & \textbf{diabetes} & Have you ever been told by a doctor or health professional that you have diabetes or sugar diabetes?\\
\textbf{BPQ020} & \textbf{hypertension} & \{Have you/Has SP\} ever been told by a doctor or other health professional that \{you/s/he\} had hypertension, also called high blood pressure?\\
\textbf{URXCRS} & \textbf{creatinine} & Creatinine, urine (umol/L)\\
\addlinespace
\textbf{SLQ040} & \textbf{snort} & In the past 12 months, how often did \{you/SP\} snort, gasp, or stop breathing while \{you were/s/he was\} asleep?\\
\textbf{LBDTCSI} & \textbf{total cholestrol levels} & Total cholesterol (mmol/L)\\
\textbf{LBXHGB} & \textbf{hemoglobin} & Hemoglobin (g/dL)\\
\textbf{LBXSASSI} & \textbf{AST} & Aspertate aminotransferase\\
\textbf{SMQ040} & \textbf{smoke} & Do you now smoke cigarettes?\\
\addlinespace
\textbf{BPXDI} & \textbf{blood pressure} & Systolic and diastolic blood pressure taken as the average of the first three measurements\\
\bottomrule
\end{tabular}
\end{table}
```
```{r tabone, echo=FALSE, eval=FALSE}
tab.one <- vars %>% 
  kable(format = "latex", caption = "Variable descriptions", booktabs = TRUE) %>% 
  kable_styling(latex_options = c("HOLD_position")) %>%
  row_spec(c(0), bold = TRUE) %>%
  column_spec(c(1, 2), bold = TRUE, width = "2cm") %>%
  column_spec(3, width = "8cm")
tab.one
```

@@

Univariate regression analysis results exploring the association between
the exposures, selected potential confounding variables showed
substantial associations between BMI ($\beta = 0.95$ mmHg; 95%CI [0.60 -
1.3]; p\<0.05), TC ($\beta = 3.6$ mmHg; 95%CI [2.9 - 4.3]; p\<0.05),
albumin ($\beta = -0.48$ mmHg; 95% CI [-0.68mmHg - -0.27]; p \< 0.05),
AST ($\beta = 0.12$ mmHg; 95%CI [0.07 - 0.18]; p \<0.05), and age
(years) ($\beta = 0.49$ mmHg; 95% CI [0.45 - 0.52]; p \<0.05) and
systolic blood pressure. Workday sleep duration between 6 and 8 hours
was associated with a 4.0 mmHg (95% CI [-6.8 - -1.2]) decrease in the
average systolic blood pressure compared to 6 hours of sleep. On the
other hand, more than 8 hours of sleep was associated with a decrease of
3.1 mmHg in systolic blood compared to 6 hours of sleep. Substantial
associations were noted between citizenship, education, number of
children below five years, gender, marital status, race, smoking status,
snorting, alcohol, and income.

@@

Table \@ref(tab:multreg) displays the estimated coefficients derived from a multiple
regression analysis involving variables. We only show sleep during
Workdays, age, total cholesterol, and the interaction effect between
total cholesterol and age. Notably, the findings reveal that an increase
of one year in age is associated with a decrease in systolic
blood pressure (SBP) by an amount of 0.04 mmHg (95%CI [-0.22 - 0.13], p = 0.60). 
Conversely, an increase of one unit in total cholesterol is
linked to a significant reduction in SBP by a factor of 3.5 mmHg (95%CI [-5.2, -1.8], p = 0.01). Furthermore, the interaction
between age and total cholesterol yielded a statistically significant
coefficient estimate suggesting that as
both age and total cholesterol concurrently increase by one unit, there
is an associated elevation in SBP by a factor of 0.11 mmHg (95%CI [0.07 - 0.14], p = 0.001). Additionally,
the coefficients attributed to sleep duration bear analogous
interpretations to those observed in the univariate scenario.
Specifically, workday sleep durations ranging between 6 and 8 hours
demonstrate a significant average decrease of 3.0 mmHg (95%CI [-5.3 - -0.67],  p = 0.012) in systolic blood pressure in
comparison to the reference duration of 6 hours of sleep. Conversely,
durations exceeding 8 hours exhibit a decrease of 1.4mmHg (95%CI [-3.7- 0.92] p = 0.2]) in systolic
blood pressure as compared to the 6-hour reference duration.

@@

\begin{table}[H]
\caption{\label{tab:multreg} Sleep, Age and Cholesterol Coefficients based on Model 5. These results are adjusted for other significant potentially confounding factors based on univariate regression analysis.}
\centering
\begin{tabular}[t]{lrrr}
\toprule
  \textbf{Variable} & \textbf{Estimate} & \textbf{95\% CI}  & \textbf{p-value} \\
\midrule
\textbf{Intercept} & 102.55 &(88.42,116.69) & $<0.001$\\
\textbf{Sleep} & 0.75 & (0.50, 1.00) & 0.002 \\
$<6$hrs &  &  &  \\
$>8$hrs & -1.4 & (-3.7,0.92) & 0.2\\
$6-8$hrs & -3.0 & (-5.3,-0.67) &  0.012\\
\textbf{Age} & -0.04& (-0.22, 0.13) & 0.6 \\
\textbf{TC} & -3.5  & (-5.2, -1.8) & <0.001 \\
\textbf{TC*Age} & 0.11  & (0.07, 0.14) & <0.001 \\
\bottomrule
\end{tabular}
\end{table}
```

@@

The current study investigated the
intricate interplay between age, sleep, cholestrol levels and systolic blood
pressure adjusting for other potentially confounding covariates informed by literature. We performed comprehensive model comparisons to explore the association between age, total cholestrol levels, sleep durations and systolic blood pressure. We find based on the best model, that work day sleep durations ranging between 6 and 8 hours and more than 8 hours are associated with a decrease in systolic blood pressure relative to the 6 hour reference group. Additionally, our study reports substantial effect modification of the effect of total cholestrol levels on systolic blood pressure by age whe adjusted for selected confounders within the study population. Previous studies exploring sleep and systolic blood pressure have found similar associations [@su2022association].
